---
title: "Design tradeoffs"
author: "Hadley Wickham"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design tradeoffs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)

fail <- function() "\u274c"
pass <- function() "\u2705"
```

There are many different ways that magrittr could implement the pipe. The goal of this document is to elucidate the variations, and the various pros and cons of each approach. This document is primarily aimed at the magrittr developers (so we don't forget about important considerations), but will be of interest to anyone who wants to understand pipes better, or to create their own pipe that makes different tradeoffs

## Code transformation

There are three main options for how we might transform a pipeline in base R expressions. Here they are illustrated with `. %>% foo() %>% bar()`:

*   __Nested__

    ```{r}
    bar(foo(.))
    ```
    
*   __Eager__

    ```{r}
    . <- foo(.)
    bar(.)
    ```
    
*   __Lazy__

    ```{r}
    ...2 %<~% foo(...1)
    bar(...1)
    ```

(There is a fourth option, which uses eager evaluation, but uses a unique variable name for each stage. This has no advantages compared to the eager pipe so we will not consider it further.)

We'll first explore the desired properties we might want a pipe to possess and then see how each of the three variants does. 

### Desired properties

These are the properties that we might want a pipe to possess, roughly ordered from most important to leasy important.

*   Visibility: the visibility of the final function in the pipe should be
    preserved. This important so that pipes that end in a side-effect function
    (which generally return their first argument invisibly) do not print.
  
*   Lazy evaluation: are steps of the pipe only evaluated lazily when actually 
    needed? This is a useful property as it means that pipes can handle 
    expressions like `stop("!") %>% try()`, which generally makes pipes more 
    expressive.

*   Eager unbinding: pipes are often used with large data objects, so 
    intermediate objects in the pipeline should be unbound as soon as possible 
    so they are available for the garbage collector.

*   Single evaluation: each component of the pipe should only be evaluated once,
    i.e. `sample(10) %>% cbind(., .)` yields two columns with the same value,
    and `sample(10) %T>% print() %>% identity()` prints and returns the same
    values.

*   Minimal stack: using the pipe should add minimal additional entries to the
    call stack, so `traceback()` is maximally useful.

### Nested pipe

*   Visibility: `r pass()`

*   Lazy evaluation: `r pass()`

*   Eager clean up: `r pass()`

*   Single evaluation: `r fail()` trivial for simple pipes, but not possible for 
    pipes that use the pronoun in multiple places. The simplest rewrite doesn't 
    work because there's no gaurantee that the first argument will be evaluated
    before the second argument.

    ```{r}
    x %>% foo(., .)
    foo(. <- x, .)
    ```

*   Minimal stack: `r fail()` maximum stack depth is the length of the pipe.

### Eager pipe

*   Visibility: `r pass()`. 

    Note that the final computation must be handled differently, as the 
    following transformation loses visibility.
    
    ```{r}
    . <- foo(.)
    . <- bar(.)
    .
    ```

*   Lazy evaluation: `r fail()` assignment forces eager evaluation of each step.

*   Eager clean up: `r pass()`

*   Single evaluation: `r pass()`

*   Minimal stack: `r pass()` maximum stack depth is 1.

### Lazy pipe

```{r}
library(rlang)

`%<~%` <- function(name, value, env = caller_env()) {
  env_bind_exprs(env, .eval_env = env, !!ensym(name) := !!enexpr(value))
}
```

*   Visibility: `r pass()`

*   Lazy evaluation: `r pass()`

*   Eager clean up: can be preserved by inserting a function call after each
    lazy assignment:
    
    ```{r}
    ...2 %<~% foo(...1)
    delayed_cleanup()
    bar(...1)
    ```
    
    `delayed_cleanup()` would be a C function that iterates through all 
    bindings in an environment, deleting any promises that have already been
    forced.

*   Single evaluation: `r pass()` by property of promises.

*   Minimal stack: `r pass()` maximum stack depth is 1.

## Execution environment

There are two possible scopes that we could evaluate the pipe in.

### Tradeoffs

No left overs.

In what scope should components of the pipe be evaluated? We can illustrate the options with two functions:

```{r}
f <- function() {
  10 %>% return()
  return(20)
}

g <- function() {
  x <- 20
  10 %>% assign("x", .)
  x
}
```

There are two scopes in which the pipe could be evaluated:

*   In a new environment (as if we had created an anonymous function and
    evaluated it), where both `f()` and `g()` return 20.
    
*   In the current environment, where both `f()` and `g()` return 10.

### Current environment

### Closure environment

